---
title: "ETC3250/5250 Tutorial 10 Instructions"
subtitle: "Model choice, and regularisation"
author: "prepared by Professor Di Cook"
date: "Week 10"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---


```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)

library(knitr)
library(tidyverse); ggplot2::theme_set(theme_bw())
library(ISLR)
library(skimr)
library(leaps)
library(patchwork)
library(rsample)
library(tidymodels)
library(glmnet)
```

## `r emo::ji("target")` Objective

### 1. Choosing variables

Work your way through the textbook lab 6.5.1 Best Subset Selection, and the forward stepwise procedure in 6.5.2, then answer these questions.

      read in data + skimr::skim
```{r}
# read in data ISLR::Hitters
data(Hitters)
```

      skim data + handle missing data on `Salary`; by removing them
```{r eval=TRUE}
# skim data
skimr::skim(Hitters)

# handle missing values on `Salary`; by removing them
Hitters <- Hitters %>%
  filter(!is.na(Salary))
```
      
      run subset selection
```{r}
# --- run subset selection
regfit.full <- leaps::regsubsets(Salary ~ .,
                                 data = Hitters,
                                 nvmax = 8, # default
                                 method = "exhaustive") # fit (p = 19 choose k = 8) models!

summary(regfit.full) # best model when k = 1, k = 2, ..., k = 8
```

      compute no. of models fitted!
•at each `k`; we fit $p \choose k$ linear models!
```{r}
# total no. of linear models fitted models = 169, 765!
total <- 0 

for(k in 1:8){
  k_no_models <- choose(n = 19,
                        k = k)
  
  total <- total + k_no_models
}
```


#### a. The `regsubsets()` function (part of the `leaps` library) performs best subset selection by identifying the best model that contains a given number of predictors, where best is quantified using RSS. By default it only examines up to 8 variable models. Which variables make the best 8 variable model?

      extract coefficients of model; where k(no. of predictors) = 8 
```{r}
regfit.full %>% 
  stats::coef(8)
```

#### b. Set the max number of variables to be 19, by adding the argument `nvmax=19`. Plot the model fit diagnostics for the best model of each size. What would these diagnostics suggest about an appropriate choice of models? Do your results compare with the text book results? Why not?

      repeat best subset selection algorithm as per above; for up to k = 19 variables
```{r}
regfit.full <- leaps::regsubsets(Salary ~ .,
                                 data = Hitters, 
                                 nvmax = 19, # up to k = 19
                                 method = "exhaustive") # fit (p = 19 choose k = 19) models!


reg.summary <- summary(regfit.full)
```
 
      
      compute no. of models fitted!
•at each `k`; we fit $p \choose k$ linear models!
```{r}
# total no. of linear models fitted models = 524, 287! 
total <- 0 

for(k in 1:19){
  k_no_models <- choose(n = 19,
                        k = k)
  
  total <- total + k_no_models
}
```
 
      what model diagnostics available?
```{r}
reg.summary <- summary(regfit.full) 

names(reg.summary)
```

      extract & plot model fit statistics; for best model for each k from 1 to 19
•note: cannot compare RSS or R^2; since they will decrease; if k increase
```{r}
models <- tibble(nvars = 1:19, # k
                 # extract model fit diagnostics; for best fit at each k from 1 to 19
                 rsq = reg.summary$rsq, # r-squared 
                 rss = reg.summary$rss, # rss
                 adjr2 = reg.summary$adjr2, # adj. r-squared
                 cp = reg.summary$cp, # Mallows C_p
                 bic = reg.summary$bic) # BIC

# --- plot model fit diagnostics against k (no. of vars.)
p1 <- models %>% 
  ggplot(aes(x = nvars, 
             y = rsq)) +
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise rsq)
  geom_point(colour = "red",
             data = models %>% slice_max(rsq, n = 1))

p2 <- models %>% 
  ggplot(aes(x = nvars, y = rss)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise rss)
  geom_point(colour = "red",
             data = models %>% slice_min(rss, n = 1))

p3 <- models %>% 
  ggplot(aes(x = nvars, 
             y = adjr2)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (maximise adjr2)
  geom_point(colour = "red",
             data = models %>% slice_max(adjr2, n = 1))

p4 <- models %>% 
  ggplot(aes(x = nvars, 
             y = cp)) +
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise cp)
  geom_point(colour = "red",
             data = models %>% slice_min(cp, n = 1))

p5 <- models %>% 
  ggplot(aes(x = nvars,
             y = bic)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise bic)
  geom_point(colour = "red",
             data = models %>% slice_min(bic, n = 1))

require(patchwork)
p1 + p2 + p3 + 
  p4 + p5
```

**BIC would suggest 6 variables. (It gets worse after 6, and then better at 8, and then worse again.) The others suggest around 10. The textbook suggests 6 variables, so similar results here.**

#### c. Fit forward stepwise selection. How would the decision about best model change?  

      fit forward stepwise selection
```{r}
regfit.fwd <- regsubsets(Salary ~ .,
                         data = Hitters, 
                         nvmax = 19, # up to k = 19
                         method = "forward") # forward stepwise selection

summary(regfit.fwd) 
```

      no. of models fitted by forward stepwise selection
```{r}
sum(1:19) # only 190!
```
      
      plot rss against k (no. of vars.)
•use elbow method
```{r}
d <- tibble(nvars = 1:19,  
            rss = summary(regfit.fwd)$rss) # extract rss

# --- plot rss against nvars
d %>% 
  ggplot(aes(x = nvars, 
             y = rss)) + 
  geom_line() +
  geom_point()
```

      extract best model at k = 6 from subset selection
```{r}
regfit.full %>% coef(id = 6)
```

      extract best model at k = 6 from forward stepwise selection
```{r}
regfit.fwd %>% coef(id = 6)
```

**We are using RSS, because this is returned by the forward stepwise procedure. Look for decreasing values, and when it flattens out. The suggestion is at 6 or 10 variables.** 

**The models with 6 predictors are identical.**
**- even though; forward stepwise seleciton; iterate over fewer models**


      can plot model diagnostics; as per above
•get very similar plots
```{r}
regfit_fwd_summary <- summary(regfit.fwd)

models <- tibble(nvars = 1:19, # k
                 # extract model fit diagnostics; for best fit at each k from 1 to 19
                 rsq = regfit_fwd_summary$rsq, # r-squared 
                 rss = regfit_fwd_summary$rss, # rss
                 adjr2 = regfit_fwd_summary$adjr2, # adj. r-squared
                 cp = regfit_fwd_summary$cp, # Mallows C_p
                 bic = regfit_fwd_summary$bic) # BIC

# --- plot model fit diagnostics against k (no. of vars.)
p1 <- models %>% 
  ggplot(aes(x = nvars, 
             y = rsq)) +
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise rsq)
  geom_point(colour = "red",
             data = models %>% slice_max(rsq, n = 1))

p2 <- models %>% 
  ggplot(aes(x = nvars, y = rss)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise rss)
  geom_point(colour = "red",
             data = models %>% slice_min(rss, n = 1))

p3 <- models %>% 
  ggplot(aes(x = nvars, 
             y = adjr2)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (maximise adjr2)
  geom_point(colour = "red",
             data = models %>% slice_max(adjr2, n = 1))

p4 <- models %>% 
  ggplot(aes(x = nvars, 
             y = cp)) +
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise cp)
  geom_point(colour = "red",
             data = models %>% slice_min(cp, n = 1))

p5 <- models %>% 
  ggplot(aes(x = nvars,
             y = bic)) + 
  geom_line() +
  geom_point() +
  # colour optimal pt. red (minimise bic)
  geom_point(colour = "red",
             data = models %>% slice_min(bic, n = 1))

require(patchwork)
p1 + p2 + p3 + 
  p4 + p5
```



### 2. Training and testing sets with variable selection

#### a. Break the data into a 2/3 training and 1/3 test set.

      split data into 2/3 training & 1/3 test set
```{r}
set.seed(1)

split <- rsample::initial_split(Hitters,
                                2/3) # 2/3 training set

# extract training & test set
h_tr <- rsample::training(split)
h_ts <- rsample::testing(split)
```
      

#### b. Fit the best subsets. Compute the mean square error for the test set. Which model would it suggest? Is the subset of models similar to produced on the full data set? Do your results compare with the text book results? Why not?

      run best subset selection
```{r}
regfit.best <- leaps::regsubsets(Salary ~ .,
                                 data = h_tr, 
                                 nvmax = 19, # up to k = 19
                                 method = "exhaustive") # best subset selection

summary(regfit.best)
```

      compute test MSE; for each of 19 best model *manually; via matrix multiplication & for loop
```{r}
# turn test set into matrix (to do matrix multiplication with \beta coefs.)
# also convert categorical vars.; into dummy vars. (a model matrix)
test.mat <- stats::model.matrix(Salary ~ ., 
                                data = h_ts)

val.errors <- rep(NA, times = 19) # storage vector; to store MSE for each of k = 19 models

for (i in 1:19) { # 1 iteration; for each model
   coefi <- regfit.best %>% 
     coef(id = i) # extract model coefs.; for ith model; where i = 1, 2, ..., 19
   
   pred <- test.mat[,names(coefi)] %*% coefi # \hat{y} = multiply \beta coef. with variables (x); get predictions for obs.
   
   val.errors[i] <- mean((h_ts$Salary - pred)^2) # find MSE (y - \hat{y})^2; append to `val.errors`
}

val.errors # MSE for k = 1, 2, ..., 19 models
```

      -> e.g. of 2nd iteration; where k = 2
```{r}
coefi <- regfit.best %>% # extract beta coef. for best model when k = 2
  coef(id = 2) # for k = 2 (model with 2 variables)

# extract `x` values; given variables
test.mat[, names(coefi)] %*% coefi # x %*% \beta
```
      
      
      plot MSE against k = 1, 2, ..., 19
```{r}
d2 <- tibble(nvars = 1:19, # k = 1, 2, ..., 19
             err = val.errors) # MSE

d2 %>% 
  ggplot(aes(x = nvars, 
             y = err)) +
  geom_line() +
  geom_point() +
  # colour optimal pt. (minimise MSE)
  geom_point(colour = "red",
             data = d2 %>% slice_min(err, n = 1)) +
  labs(y = "err (MSE)")

# extract k that minimises MSE
d2 %>% 
    filter(err == min(err))
```

      extract coefs. where $k$ minimise MSE
```{r}
regfit.best %>% 
  coef(id = which.min(val.errors)) # extract; where index is minimum  *k = 3
```
      
      extract coefs for 3 predictors; from when we fit best subset selection; on FULL data
```{r}
regfit.full %>% 
  coef(id = 3)
```

**The model selection of best subsets provides different results, 10 predictors, which is likely due to using a subset of data for training the model. Note that we used the test error as the measure for choosing models, and the different metric could produce different results.  (Interestingly, if a different random seed is used, you might get a different best model. In this case, you should select the variables that consistently get used in the best model from different seeds)**

**The selected variables are the same as the 10 variable model fitted to the full set. The coefficients in the fitted model differ a little.**

> i.e. use cross validation; since changing seed; result in different best model; due to different subset of training & test split
> K-fold cv; ensure; best model that most consistently appears; across the various folds

### 3. Cross-validation with variable selection

It is said that 10-fold cross-validation is a reasonable choice for dividing the data. What size data sets would this create for this data? Argue whether this is good or bad. 

**There isn't a lot of data. With 10-fold cross-validation only about 20 cases are kept out each time, which leads to substantial variability between predictions from each set. I would suggest using 5-fold. (Note from running CV: When I run this it also produces results with considerable variability. The choice of number of variables is consistent for most $k$, even though the variability in error is substantial.)**


```{r eval=FALSE, echo=FALSE}
# --- code to do cross-validation, if you are interested

# --- create own prediction function *a method: for running `predict` on `regsubsets` objects & each fold
predict.regsubsets <- function(object, newdata, id, ...) {
  # extract fold
  form <- as.formula(object$call[[2]])

  # create model matrix; for test set
  mat <- model.matrix(form,
                      newdata)

  # X: extract model coefs.
  coefi <- coef(object,
                id = id)

  # \beta: extract predictors
  xvars <- names(coefi)

  # X %*% \beta
  mat[, xvars] %*% coefi
}

# --- create folds; via `caret` package
set.seed(20190513)
k <- 5
folds <- caret::createFolds(Hitters$Salary,
                            k = k)

# --- compute errors for different folds; manually; via (nested) for loops
cv.errors <- matrix(NA, k, 19, dimnames = list(NULL, paste(1:19))) # create storage matrix

for (j in 1:k) { # for 1:k folds
  # run best subset selection
  best.fit <- leaps::regsubsets(Salary ~ .,
                                data = Hitters[(1:263)[folds[[j]]], ], 
                                nvmax = 19)
  
  for (i in 1:19) {
    # compute the cv error on test set; for each of the k = 1, 2, ..., 19 models
    pred <- predict(best.fit,
                    Hitters[folds[[j]],],
                    id = i)
    
    cv.errors[j, i] <- mean((Hitters$Salary[folds[[j]]] - pred)^2) # compute cross validation error; for jth fold & ith model
  }
}

# --- put results in `tibble`; in long form for plotting
# cv.errors <- as_tibble(cv.errors) %>% # rows = test error// column = k (no. of variables)
#   gather(nvars, error) %>% # superseded by pivot_longer (more intuitive to use)
#   mutate(nvars = as.numeric(nvars)) %>%
#   mutate(cv_set = rep(1:5, 19))

# OR using `pivot_longer`
cv.errors <- as_tibble(cv.errors) %>% 
  pivot_longer(cols = everything(),
               names_to = "nvars",
               values_to = "error") %>% 
  mutate(nvars = as.numeric(nvars)) %>% 
  mutate(cv_set = rep(1:5, each = 19)) # 19 models for first fold// 19 models for 2nd fold// ...// 19 models for 5th fold

# --- plot results
cv.errors %>% 
  ggplot(aes(x = nvars,
             y = error,
             colour = factor(cv_set))) +
  geom_line() + 
  theme(legend.position = "none")
```

### 4. Regularisation for variable selection

Here we will use lasso to fit a regularised regression model and compare the results with the best subset model.

#### a. Using your results from questions 1-3, fit the best least squares model, to your training set. Write down the mean square error and estimates for the final model. We'll use these to compare with the lasso fit.

```{r}
min(val.errors)
coef(regfit.best, which.min(val.errors))
```

#### b. Fit the lasso to a range of $\lambda$ values. Plot the standardised coefficients against $\lambda$. What does this suggest about the predictors?

```{r}
grid <- 10^seq(10,-2,length=100)
x <- model.matrix(Salary~., h_tr)[,-1]
y <- h_tr$Salary

lasso.mod <- glmnet(x, y, alpha=1, lambda=grid)
# Need a coefficient matrix of 100(nlambda)x19(p)
# as.matrix function converts sparse format into this
coeffs <- as.matrix(lasso.mod$beta) 
coeffs <- cbind(var=rownames(coeffs), coeffs)
cv <- coeffs %>% as_tibble() %>%
  gather(nval, coeffs, -var) %>%
  mutate(coeffs=as.numeric(coeffs)) %>%
  mutate(lambda=rep(lasso.mod$lambda, rep(19, 100)))
  
p <- ggplot(cv, aes(x=lambda, y=coeffs, group=var, label=var)) + geom_line() +
  scale_x_log10(limits=c(-1, 100))
plotly::ggplotly(p)
# This is how the sample code plots
#plot(lasso.mod, xvar="lambda", xlim=c(-1, 5))
```

**As seen from the few lines that are not near zero, there are just a few predictors that are important for predicting salary.**

#### c. Now use cross-validation to choose the best $\lambda$.

```{r}
# Do cross-validation, using glmnet's function
set.seed(1)
cv.out <- cv.glmnet(x, y, alpha=1)
#cv.df <- tibble(lambda=cv.out$lambda, mse=cv.out$cvm,
#                mse_l=cv.out$cvlo, mse_h=cv.out$cvup)
#ggplot(cv.df, aes(x=lambda, y=mse)) + geom_point() +
#  scale_x_log10() +
#  geom_errorbar(aes(ymin=mse_l, ymax=mse_h))
# This is how the sample code plots
plot(cv.out)
```

#### d. Fit the final model using the best $\lambda$. What are the estimated coefficients? What predictors contribute to the model?

```{r}
# Fit a single model to the best lambda, predict the test set, and compute mse
bestlam <- cv.out$lambda.min
bestlam
x_ts <- model.matrix(Salary~., h_ts)[,-1]
y_ts <- h_ts$Salary
lasso.pred <- predict(lasso.mod, s=bestlam, newx=x_ts)
y.test <- y_ts
mean((lasso.pred-y.test)^2)

# Fit the model
# alpha=1 means lasso
# Its strange that it still needs the grid of lambdas to fit
# but it seems necessary for the optimisation.
# Then the predictions for best lambda made with predict fn
out <- glmnet(x, y, alpha=1, lambda=grid)
lasso.coef <- predict(out, type="coefficients", s=bestlam)[1:20,]
lasso.coef
lasso.coef[lasso.coef!=0]
```

**With the seed provided there are 13 non-zero coefficients, and an MSE of 67287.59.**

#### e. Does the best lasso model beat the best least squares model (best subsets)? 

**The best subsets performs a little better than lasso. It has lower MSE. The lasso has fewer variables, though, and thus is a little easier to interpret.**

### 5. Making sense of it all

Only one variable is very important for the model, which is it? (It occurs in every list of variables returned as important, and has the highest coefficient in the lasso model.) Several more variables frequently appear as important, which ones are these? Several others, appear occasionally in some models but always have very small coefficients. Can you name one of these? What does this tell you about the relationship between Salary and all the predictors?

**DivisionW is by far the most important variable. It is always selected, and always has a large coefficient.**

**AtBat, Hits, Walks persistently appear with relatively small coefficients, and appear in the lasso.**

**PutOuts, CBRI and Assists appear regularly with really small coefficients.**

**In the lasso model, it is interesting that the variable NewLeagueN appears to be important, but it is the variable who's coefficient is reduced to 0 quickly. It also never shows up in any of the subset selection models. Also, years makes an appearance in the lasso model, and is included as a non-zero coefficient in the final model, which differs from all the subset selection models.**

**There is only one major variable useful for predicting salary, which is DivisionW. This variable alone provides most of the predictive power. Small gains are obtained by adding additional variables.**

##### © Copyright 2021 Monash University
