---
title: "ETC3250/5250 Tutorial 5 Solution"
subtitle: "Visualising high-dimensional data"
author: "prepared by Professor Di Cook"
date: "Week 5"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)

# --- load libraries
library(tidyverse); ggplot2::theme_set(theme_bw())
library(discrim)
library(emo)
library(htmltools)
library(MASS)
library(parsnip)
library(plotly)
library(rsample)
library(spinifex)
library(tourr)
library(yardstick)
```



## 💬 Class discussion exercises

- In tour 1, do you see any clustering? 

      read in data
```{r eval=FALSE, echo=TRUE}
# --- read in data
olive <- read_csv("http://www.ggobi.org/book/data/olive.csv") %>%
  dplyr::select(-`...1`)
```
      
      grand tour 1 (w/o colouring response)
•$p = 8// d = 2$
```{r}
# --- run grand tour (simple)
tourr::animate_xy(data = olive[, 3:10],
                  tour_path = grand_tour(d = 2),
                  axes="off") # turn off axes (loading scores)

# --- run grand tour (customised)
set.seed(20190331)

bases <- tourr::save_history(data = olive[,3:10], 
                             tour_path = grand_tour(d = 2), 
                             start = matrix(c(1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0),
                                            ncol = 2,
                                            byrow = TRUE), 
                             max = 15)

# Re-set start bc seems to go awry
bases[,,1] <- matrix(c(1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0), 
                     ncol = 2,
                     byrow = TRUE)

tour_path <- tourr::interpolate(basis_set = bases,
                                0.1)

d <- dim(tour_path)

olive_std <- tourr::rescale(olive[,3:10])

mydat <- NULL; 
for (i in 1:d[3]) {
  fp <- as.matrix(olive_std) %*% matrix(tour_path[,,i], ncol=2)
  fp <- tourr::center(fp)
  colnames(fp) <- c("d1", "d2")
  mydat <- rbind(mydat, cbind(fp, rep(i+10, nrow(fp))))
}

colnames(mydat)[3] <- "indx"

df <- as_tibble(mydat) 

p <- ggplot() +
       geom_point(data = df, aes(x = d1, 
                                 y = d2, 
                                 frame = indx), 
                  size=1) +
  theme_void() +
  coord_fixed() +
  theme(legend.position="none")

pg <- p %>% 
  plotly::ggplotly(width = 400, 
                   height = 400) %>%
  plotly::animation_opts(frame = 200, 
                         redraw = FALSE, 
                         easing = "linear", 
                         transition = 0)

htmltools::save_html(pg, 
                     file = "tour1.html")
```

      
- In tour 2, where three classes have been coloured, how many additional clusters do you see? [tour2](https://iml.numbat.space/labs/tour2.html)

- In tour 2, the three classes have been coloured. can you see differences between the three groups? Do any of these group break into more clusters?

      grand tour 2; map response(`region`) to colour aes
```{r eval=FALSE, echo=TRUE}
# This is code that YOU CAN RUN YOURSELF to see the tour, 
# but its not necessary to run in order to do the exercise 

# --- run grand tour (simple)
  tourr::animate_xy(data = olive[,3:10], # the 8 fatty acid variables
                  tour_path = grand_tour(d = 2),
                  axes = "off", # turn axes off
                  col = olive$region)

# --- run grand tour (customised)
df <- df %>%
  mutate(region = factor(rep(olive$region, d[3])))

p <- ggplot() +
  geom_point(aes(x = d1, 
                 y = d2, 
                 colour = region,
                 frame = indx), 
             size = 1,
             data = df) +
  scale_colour_brewer("", palette="Dark2") +
  theme_void() +
  coord_fixed() +
  theme(legend.position="none")

pg <- p %>% 
  plotly::ggplotly(width = 400,
                   height = 400) %>%
  plotly::animation_opts(frame = 200, 
                         redraw = FALSE, 
                         easing = "linear", 
                         transition = 0)

htmltools::save_html(pg, 
                     file = "tour2.html")
```

      extra: guided tour LDA PP
•find projection; where 3 regions; most separated
```{r}
tourr::animate_xy(data = olive[,3:10], # the 8 fatty acid variables
                  tour_path = guided_tour(lda_pp(olive$region)),
                  axes = "off", # turn axes off
                  col = olive$region)
```
      

# `r emo::ji("gear")` Exercises 

## 1. Data with different variance-covariances

Take a look at the data from tutorial 5 using a grand tour. Can you see the difference between the two sets better now? You should see that one group has roughly the same spread of observations for each group, in any combination of the variables. The other has some combinations of the variables where the spread is different for each group. 

**Set A** has equal variance-covariance between groups, $\Sigma$: 

$$\Sigma = \begin{bmatrix} 3.0&0.2&-1.2&0.9\\
0.2&2.5&-1.4&0.3\\
-1.2&-1.4&2.0&1.0\\
0.9&0.3&1.0&3.0\\
\end{bmatrix}$$
 
and **set B** has different variance-covariances between groups, $\Sigma_1, \Sigma_2, \Sigma_3$:

$\Sigma_1 = \Sigma$

$$\Sigma_2 = \begin{bmatrix}3.0&-0.8&1.2&0.3\\
-0.8&2.5&1.4&0.3\\
1.2&1.4&2.0&1.0\\
0.3&0.3&1.0&3.0\\
\end{bmatrix}$$

$$\Sigma_3 = \begin{bmatrix}2.0&-1.0&1.2&0.3\\
-1.0&2.5&1.4&0.3\\
1.2&1.4&4.0&-1.2\\
0.3&0.3&-1.2&3.0\\
\end{bmatrix}$$

Make a scatterplot matrix to match the variance-covariance matrix with the spread of the observations. In setA, the spread is (approximately) **homogeneous** between groups. In setB, the spread is clearly **heterogeneous** between groups. 

      code to simulate multivariate normal data
```{r echo=TRUE, eval=FALSE}
set.seed(20200416)
library(mvtnorm)

# --- variance-covariance matrix
vc1 <- matrix(c(3, 0.2, -1.2, 0.9, 0.2, 2.5, -1.4, 0.3, -1.2, -1.4, 2.0, 1.0, 0.9, 0.3, 1.0, 3.0), 
              ncol = 4, 
              byrow = TRUE) # create matrix by row

vc2 <- matrix(c(3, -0.8, 1.2, 0.3, -0.8, 2.5, 1.4, 0.3, 1.2, 1.4, 2.0, 1.0, 0.3, 0.3, 1.0, 3.0), 
              ncol = 4, 
              byrow = TRUE)

vc3 <- matrix(c(2.0, -1.0, 1.2, 0.3, -1.0, 2.5, 1.4, 0.3, 1.2, 1.4, 4.0, -1.2, 0.3, 0.3, -1.2, 3.0),
              ncol = 4,
              byrow = TRUE)

# --- mean of each group
m1 <- c(0,0,3,0)

m2 <- c(0,3,-3,0)

m3 <- c(-3,0,3,3)

# no. of obs. in each group
n1 <- 85

n2 <- 104

n3 <- 49

# === simulate multivariate normal density 
# --- set A
setA <- rbind(mvtnorm::rmvnorm(n = n1, mean = m1, sigma = vc1), # group 1
              mvtnorm::rmvnorm(n = n2, mean = m2, sigma = vc1), # group 2
              mvtnorm::rmvnorm(n = n3, mean = m3, sigma = vc1)) %>% # group 3
  data.frame()

# add `class` column
setA$class <- c(rep("1", n1), 
                rep("2", n2), 
                rep("3", n3))

# --- set B
setB <- rbind(mvtnorm::rmvnorm(n = n1, mean = m1, sigma = vc1), # group 1
              mvtnorm::rmvnorm(n = n2, mean = m2, sigma = vc2), # group 2
              mvtnorm::rmvnorm(n = n3, mean = m3, sigma = vc3)) %>%  # group 3
  data.frame()

# add `class` column
setB$class <- c(rep("1", n1), 
                rep("2", n2), 
                rep("3", n3))
```


      make a scatterplot matrix
•
```{r eval=FALSE}
# --- for set A
setA %>% 
  GGally::ggscatmat(columns = 1:4,
                    color = "class") 

# OR

# setA %>% 
#   GGally::ggpairs(columns = 1:4,
#                   ggplot2::aes(colour = class)) # colour by `class`

# --- for set B
setB %>% 
  GGally::ggscatmat(columns = 1:4,
                    color="class")

# OR

# setB %>% 
#   GGally::ggpairs(columns = 1:4,
#                   ggplot2::aes(colour = class)) # colour by `class`
```

You can use this code to run the tour (RUN CODE ONE LINE AT A TIME!):


```{r eval=FALSE}
tourr::animate_xy(data = setA[,1:3],
                  tour_path = grand_tour(d = 2),
                  col = setA$class)

tourr::animate_xy(data = setB[,1:4],
                  tour_path = grand_tour(d = 2),
                  col = setB$class)
```

Also note that if you include the fourth variable `1:4` in the code above, that the variance-covariance of setA is collinear (all the points align along a line) in some projections. That means that the spread or variation in each group is only 3D. This is something that a tour can help you see that was not at all visible from the scatterplot matrix. It can also be determined from PCA, which would return a 0 eigenvalue for the fourth PC, and the first three PCs would togetehr explain 100% of the total variance. 
**You should be able to see that set B has two groups where the variance-covariance is elliptical in shape but oriented very differently.**


## 2. Exploring for class separations, heterogeneous variance-covariance and outliers

Remember the chocolates data? The chocolates data was compiled by students in a previous class of Prof Cook, by collecting nutrition information on the chocolates as listed on their internet sites. All numbers were normalised to be equivalent to a 100g serving. Units of measurement are listed in the variable name. You are interested in answering "How do milk and dark chocolates differ on nutritional values?"

### a. Examine all of the nutritional variables, relative to the chocolate type, using a grand tour (`tourr::animate_xy()`) and a guided tour (look up the help for `tourr::guided_tour` to see example of how to use the `lda_pp` index). Explain what you see in terms of differences between groups.

      read in data + standardise variables
```{r echo=TRUE, eval=FALSE}
# --- read in data
choc <- read_csv("http://iml.numbat.space/data/chocolates.csv")

# rename variables; to shorter names
choc <- choc %>%
  rename(Cl = Calories, 
         CF = CalFat,
         TF = TotFat_g,
         SF = SatFat_g,
         Ch = Chol_mg,
         Na = Na_mg,
         Cb = Carbs_g,
         Fb = Fiber_g,
         Sg = Sugars_g,
         Pr = Protein_g)

# standardise variables
std <- function(x) (x - mean(x)) / sd(x) # create function to standardise vars.

choc_std <- choc %>%
  mutate_if(is.numeric, std)

# OR

# choc_std <- choc %>% 
#   mutate(across(.cols = where(is.numeric),
#                 .fns = std))
```

      run grand tour + guided tour 
```{r}
# --- run grand tour
tourr::animate_xy(data = choc_std[,5:14], # select all numeric vars.
                  tour_path = grand_tour(d = 2),
                  col = choc$Type) # colour pts. by `Type`

# --- run guided tour; LDA pp
tourr::animate_xy(data = choc_std[,5:14], # select all numeric vars.
                  tour_path = guided_tour(lda_pp(choc_std$Type)), 
                  col = choc_std$Type) # colour pts. by `Type`
```

**You can see some differences between the two groups, that they are almost separated, and some strong linear associations, in some combinations of variables.** 

**The guided tour with LDA index stops at a projection where most milk chocolates are separated from most dark chocolates, although there is no gap between the two clusters.**

### b. From the tour, should you assume that the variance-covariance matrices of the two types of chocolates is the same? Regardless of your answer, conduct a linear discriminant analysis, on the standardised chocolates data. Because the variables are standardised the magnitude of the coefficients of the linear discriminant can be used to determine the most important variables. What are the three most important variables? What are the four least important variables? Look at the data with a grand tour of only the three important variables, and discuss the differences between the groups. 

      fit LDA
```{r}
# --- conduct LDA
choc_std <- choc_std %>%
  mutate(Type = as.factor(Type)) # turn response (`Type`) into factor variable

# --- model specification with `parsnip`
lda_mod <- parsnip::discrim_linear() %>% # (I) specify model
  # (II) set engine
  parsnip::set_engine("MASS", 
                      prior = c(0.5, 0.5)) %>% 
  parsnip::set_mode(mode = "classification") # (III) set mode

lda_mod %>% 
  translate()

# fit LDA model
choc_lda_fit <- lda_mod %>% 
  fit(Type ~ ., 
      data = choc_std[, 4:14]) # all numeric variables + response `Type`
```

```{r}
tourr::animate_xy(choc_std[,c(7,10,12)],
                  tour_path = grand_tour(d = 2),
                  col=choc$Type)
```


**The variance-covariance matrices are clearly different between the two groups.**

**The three most important variables are Fiber_g, Na_mg and TotFat_g**

**The least important variables are Calories, CalFat, SatFat_g, Carbs_g.**

**Using just the three most important variables in the tour, you can see the differences between the two groups more clearly.**




## 3. Assessing variable importance with the manual tour

This example uses the olive oils data. 

a. Read in the data. Keep `region` and the fatty acid content variables. Standardize the variables to have mean and variance 1.
b. Fit a linear discriminant analysis model, to a training set of data. This will produce a 2D discriminant space because there are three variables. Based on the coefficients which variable(s) are important for the first direction, and which are important for the second direction?

**eicosenoic is very important for the first direction, and oleic and palmitic are most important for the second direction.**

c. Using a manual tour, with the `play_manual_tour` function from the `spinifex` package, , starting from the projection given by the discriminant space explore the importance of (i) `eicosenoic` for separating region 1, (ii) `oleic` and `linoleic` for separating regions 2 and 3, (iii) and that `stearic` is not important for any of the separations.

**(i) When eicosenoic is removed from the projection, region 1 observations get mixed/confused with the other regions.**

**(ii) When oleic is removed regions 2 and 3 get confused, BUT when oleic is 100% in the projection, the separation is also a little worse. This means that a combination of oleic and palmitic is better for separating the two groups. It is even more interesting when linoleic contributes more to the projection because the separation between groups is even better. LDA doesn't use gaps between clusters to build the separation, it uses the groups means, and makes the assumption of equal variance-covariance, so it misses this improved solution.**

**(iii) The separation between all groups is better when stearic does NOT contribute to the projection.**

      read in data + standardise variables
```{r echo=TRUE, eval=FALSE}

# --- read in data
olive <- read_csv("http://ggobi.org/book/data/olive.csv") %>%
  # dplyr::filter(region != 1) %>%
  # dplyr::select(region, oleic, linoleic) %>%
  dplyr::select(-`...1`, -area) %>%
  mutate(region = factor(region))

# --- standardise variables
std <- function(x) (x - mean(x)) / sd(x)

olive_std <- olive %>%
  mutate_if(is.numeric, std)
```
      
      split data into training & test
```{r}
set.seed(775)

olive_split <- rsample::initial_split(olive_std,
                                      prop = 2/3, 
                                      strata = region) # ensure; training & test; have same proportion of `region`

olive_train <- rsample::analysis(olive_split) # extract training set
olive_test <- rsample::assessment(olive_split) # extract test set
```

```{r}
# --- model specification with `parsnip`
lda_mod <- parsnip::discrim_linear() %>% # (I) specify model
  # (II) set engine
  parsnip::set_engine("MASS", 
                      prior = c(1/3, 1/3, 1/3)) %>% 
  parsnip::set_mode(mode = "classification") # (III) set mode

lda_mod %>% 
  translate()

# fit LDA model to training set
olive_lda_fit <- lda_mod %>% 
  fit(region ~ ., 
      data = olive_train)
```

      manual tour; starting from projection given by discriminant space; examine importance of `eicosenoic`
```{r}
# collect the projection giving best separation from LDA
# to use with manual tour for assessing importance
lda_proj <- olive_lda_fit$fit$scaling

# generate the manual path; examining the importance of eicosenoic (variable 8)
path1 <- spinifex::manual_tour(basis = lda_proj, 
                               manip_var = 8) # <<

ggt <- spinifex::ggtour(basis_array = path1, 
                        data = olive_std[,2:9],
                        angle = 0.05) + 
  # adds `geom_point` of projected data
  spinifex::proto_point(aes_args = list(color = olive_std$region)) + 
  proto_basis()

spinifex::animate_plotly(ggt)
```

      manual tour; examine importance of `oleic`
```{r}
# === generate the manual path examining the importance of oleic (variable 4) - Change the 8 to a 4 in the previous code

# generate the manual path; examining the importance of `oleic` (variable 4)
path1 <- spinifex::manual_tour(basis = lda_proj, 
                               manip_var = 4) # <<

ggt <- spinifex::ggtour(basis_array = path1, 
                        data = olive_std[,2:9],
                        angle = 0.05) + 
  # adds `geom_point` of projected data
  spinifex::proto_point(aes_args = list(color = olive_std$region)) + 
  proto_basis()

spinifex::animate_plotly(ggt)
```

      manual tour; examine importance of `linoleic`
```{r}
# === Generate the manual path examining the importance of `linoleic` (variable 5) - Change the 8 to a 5 in the previous code

# generate the manual path; examining the importance of `linoleic` (variable 5)
path1 <- spinifex::manual_tour(basis = lda_proj, 
                               manip_var = 5) # <<

ggt <- spinifex::ggtour(basis_array = path1, 
                        data = olive_std[,2:9],
                        angle = 0.05) + 
  # adds `geom_point` of projected data
  spinifex::proto_point(aes_args = list(color = olive_std$region)) + 
  proto_basis()

spinifex::animate_plotly(ggt)
```

      manual tour; examine importance of `arachidic`
```{r}
# === Generate the manual path examining the importance of `arachidic` (variable 7) - Change the 8 to a 7 in the previous code 

# generate the manual path; examining the importance of `arachidic` (variable 7)
path1 <- spinifex::manual_tour(basis = lda_proj, 
                               manip_var = 7) # <<

ggt <- spinifex::ggtour(basis_array = path1, 
                        data = olive_std[,2:9],
                        angle = 0.05) + 
  # adds `geom_point` of projected data
  spinifex::proto_point(aes_args = list(color = olive_std$region)) + 
  proto_basis()

spinifex::animate_plotly(ggt)
```

      manual tour; examine importance of `stearic`
```{r}
# === Generate the manual path examining the importance of stearic (variable 3) - Change the 8 to a 3 in the previous code

# generate the manual path; examining the importance of `stearic` (variable 3)
path1 <- spinifex::manual_tour(basis = lda_proj, 
                               manip_var = 3) # <<

ggt <- spinifex::ggtour(basis_array = path1, 
                        data = olive_std[,2:9],
                        angle = 0.05) + 
  # adds `geom_point` of projected data
  spinifex::proto_point(aes_args = list(color = olive_std$region)) + 
  proto_basis()

spinifex::animate_plotly(ggt)
```


##### © Copyright 2022 Monash University
